Discovering cell type motifs in connectomics data

### 2013-12-24 docs/paper.tex
making a finite version for comparison

%\documentclass{nature}
\documentclass{article}
\usepackage[english,american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{todonotes}
\usepackage{subfiles}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}

@@

\todo{Citations}
\todo{Log score plots for real data} 
\todo{Hierarhcical structure?}
% \begin{affiliations}
%   \item Something Something
%   \item Something Something
% \end{affiliations}

\begin{abstract}
  New techniques produce massive data about neural connectivity,
  necessitating new analysis methods to discover underlying
  structure. There is strong evidence for microcircuitry, where the
  probability of connections depends on the types of pre- and
  post-synaptic neurons and their distance. Here we developed a
  nonparametric Bayesian technique that identifies neuron type and circuitry
  patterns using only this connection data. We show that the approach recovers
  known neuron types in the retina, reveals interesting structure in
  the nervous system of c. elegans, and automatically discovers the
  structure of microprocessors. Our approach is a first step towards
  condensing connectomics data into meaningful, human readable
  structure.
\end{abstract}

@@

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more compelx than traditional motifs,
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development process.

@@

\begin{figure}
  \centering 
    \centerline{\includegraphics[width=183mm]{overview.ai}}
  \caption{An overview of our method. a. Input data is
an adjacency matrix indicating whether cell $i$ (row)
synapses on cell $j$ (column). The location of each cell in space is known. 
b. Our algorithm discovers hidden cell types in this connectivity data
by assuming all cells of a type share a distance-dependent connectivity 
profile, c.), with cells of other types. This results in a clustering 
of the cells by those hidden types. b.) shows the cell adjacency
matrix with cells of the same type grouped together. d.) shows
the learned probability of connection between our different types
at various distances, illustrating short-range and long-range
connectivtiy patterns.}
\label{fig:overview}
\end{figure}

\section{Results}
We use a Bayesian structure discovery approach that is based on
assumptions about the way positions and cell types affect the
probability of connections between cells, a so-called generative
model. The probability of connections between pairs of neurons
decreases with distance, but how high the baseline probability of
connection is and how quickly it decays with distance depends on the
two cell-types. We combine annealing with Markov Chain Monte Carlo to
perform simultaneously joint posterior inference over the structure,
parameters, and hyperparameters of the model (see Methods). This
procedure yields a division of all neurons into cell-types along with
the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). These results can then be compared with the
results of previous anatomical studies that have used other aspects of
neurons, such as chemical properties and cell-morphology to divide
neurons into types.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus started out by
simulating data for which we know the correct structure and comparing
the estimated structure based on the algorithm (see methods) with the
one we used for simulation. We find that the model does a good job of
recovering the structure we used to generate the data (fig~\ref{fig:synthetic}a)
Indeed, it does far better than a simpler and well established
model (stochastic block) that assumes that there is no distance
dependence to the connectivity. The model converges relatively
quickly, which is enabled by xxx cool tricks we used to speed it
up. It appears that the model behaves the way it should and allows
application to large datasets.

\begin{figure}
  \centering 
  \centerline{\includegraphics{synthetic.ai}}
  \caption{Correct recovery of true numbers of hidden classes
in synthetic data. a.) Other models overcluster the number
of classes as they fail to take distance into account, whereas
our model finds close to the true number of classes. b.) ARI, a measure
of ``correct clustering'', for
different true class counts, between our model and a traditional
nonparametric stochastic block model. c.) Multiple Markov chains iterateively
converge towards a highly-probable solution. }
\label{fig:synthetic}
\end{figure}

@@

The algorithm took 8 hours to converge to a locally optimal solution,
dividing neurons into a set of cell types (fig~\ref{fig:retina}c, each
wedge is a type). For each pair of neurons there is a specific
distance dependent connection probability (fig~\ref{fig:retina}b,c,d),
which is well approximated by the model fit. Moreover, each type of
cell is rather isotropically distributed across space
(fig~\ref{fig:retina}e) as should be expected for true cell types. The
algorithm is thus able to extract meaningful structure from the data.

Comparing the results of the algorithm to other information sources
allows evaluating the quality of the type determination. Our types
closely reflect the (anatomist-determined) segmentation of cells into
retinal ganglion, narrow amacrine, medium/wide amacrine, and bipolar
cells (Figure 2a, outermost ring). We find that the types we find tend
to reflect the known laminar distribution in the retina
(fig~\ref{fig:retina}c, middle ring), which is exciting as the
algorithm did not actually use this type of information. The algorithm
yields a separation of neurons into a smaller number of types than the fully
granular listing of 71 types found by the original authors of the paper, 
although is still highly correlated with those finer type distinctions (see supp~\ref{supp:mouseretina}. It is our expectation that, with larger datasets, 
even closer agreement would be found.  [TODO:
Do we assign any of the unknown cells correctly?  are types 72-79
“different” or simply unknown?]

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome. 
a.) Input adajacency data for 950 cells for which soma positions were known. b. clsutered adjacency matrix. c.) connectivity diagram showing our clusters, as
well as the cell depth and known cell type of each cell. Our model agrees
with the cell types traditionally identified by anatomists, and identifies 
laminar-specific connectivity patterns. d.) Connectivity between our
clusters as a function of distance -- the cluster consisting primarially of
retinal ganglion cells (lower left-center) exhibits the expected near and
far connectivity.}
\label{fig:mouseretina}
\end{figure}

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons deterministically known. This
incredible cross-organism consistency led to early mapping of the
c. elegans chemical and electrical (gap junction) connectome. Unlike
the retina, only the motor neurons in c. elegans exhibit regular
distribution in space, being distributed along the body axis. Most
interneurons are concentrated in various ganglia that project
throughout the entire animal, and the sensory neurons are primarily
located in a small number of very anteriior ganglia.

@@

We identify cell types that reflect the known motor/non-motor neuron
classes, even though this system lacks the strong repeat
microcircuitry our model was designed for.  Motorneuron types AS, DA,
and VA, all exclusively postsynaptic, are clustered together, as are
motorneuron types VD and DD. Traditional types VC, DB, and VB also
mostly share a cluster. Various head motor neurons, including types
SMD, RMD, and RIM are clustered together. Interneurons with known
anatomically-distinct connectivity patterns, such as RIA (2 cells) or
AVA (2 cells) are clustered into pure types. The algorithm even
correctly places the single-cell type DVB by itself. Note our
clustering does not perfectly reflect known divisions -- several
combinations of head and sensory neurons are combined, and a
difficult-to-explain group of 2 VD, 2 VB, a VA, and a DD neuron are
grouped together. Our identified cell types thus reflect a
``coarsening'' of known types, based entirely on connectivity and
distance information, evne when the organism exhibits substantially
less spatial regularity than the retina.

\begin{figure}
  \centering 
  \centerline{\includegraphics{celegans.ai}}
  \caption{Discovering cell types in the \textit{c. elegans}
    connectome. a.) Initial cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportaionl to synapse count between
    those cells. b.) Clustered adjaceny matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). }
  \label{fig:celegans}
\end{figure}

@@

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segementing
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:microprocessor}a, showing the man-made horizontal and
vertical layout of the same types of transistors.

@@

  Larger datasets will generally allow algorithms to distinguish more
  distinct types (cite supplemental data) and more data should
  algorithms to get closer to the results from anatomy. Moreover, in
  general, for such problems precision increases with the size of the
  dataset and the cells that we have are not sufficient to
  statistically distinguish all the cell types known in anatomy
  (71xxx). Still, using only connectivity and distance it is possible
  to meaningfully divide neurons into types.

@@

There have existed a a range of previous approaches to the discovery
of neural microcircuitry (cites). These generally involve a great deal
of manual labor and ad-hoc determination of what constitutes a “type”
of cell -- to this day there are disagreements in the literature as to
the “true” types in the mammalian retina. Much as phylogenomics has
changed our understanding of animal ontologies, modern large scale
data will allow the efficient unbiased discovery of cell types and
circuits. The sheer amount of available data demands the introduction
of algorithmic approaches.

The development of automatic identification and quantification of cell
type may also provide new phenotypic tools for quantifying the effect
of disease, genetic abnormalities, and genetic
interventions. Particular pathologies and genetic manipulations may
subtly change the number and connectivity of cell types in ways
only visible with algorithmic analysis.

@@

We then jointly infer the maximum a posteriroi (MAP) estimate of the
class assignment vector ${c_i}$, the parameter matrix $\eta_{mn}$, and
the global model hyperparameters $\theta$ :

@@

% Methods section 3000 words
% methods summary does not appear in the online text
% methods section can have no figures or tables [grrr]
% todo describe multiple independent relations?

@@

Our model is a extension of infinite stochastic block models
\autocite{Kemp2006a,Xu2006} to incorporate spatial constraints,
inspired by attempts to extend these models with arbitrary
discriminative functions\autocite{Murphy2012}.

@@

We then employ a transition kernel to update the per-component
parameter values $\eta_{mn}$. Conditioned on the assignment vector
$\vec{c}$ and the model hyperparameters $\theta, \alpha$ the 
individual $\eta_{mn}$ are indepednent. We slice sample \autocite{Neal2003}
each component's parameters, choosing the slice width as a function
of the global hyperparameter range.

@@

To pick the MAP, we run up to $100$ simultaneous chains in parallel,
iniitalized from different random initial points in the statespace,
and then pick the chain with the highest log likelihood. Early
experiments attempted to use the full ensemble of chains as an
estimate of the posterior but were abandoned due to the difficulty of
summarizing the posterior over such a complex state space, whcih is an
ongoing area of research in the probabilistic modeling community
\autocite{}.

@@

\subsection {Mouse Retina}
Dense serial electron microscopy of a $VOLUMNE$ in the mouse retina by
\autocite{Helmstaedter2013} yielded a listing of places where neurons
come into contact. There were over 1000 cells originally, and selected
the $950$ for which the location of the soma could be reconstruted
from the provided cell plots (soma locations were not provided by the
study's authors in machine-readble form). Ultimately
this left a matrix between the total synapse-like contact area between
all pairs of 950 cells. Area was thresholded at $0.5\mu m$, determined
by hand, to yield a 950 $\times$ 950 entry matrix that served as input
to our algorithm. We measured the distance between cells using
the reconstructed soma centers, and used the Logistic-Distance
spatial relation. Hyperprior griddings are shown in \ref{supplemental}.

@@

We obtained the connectome of c. elegans from \autocite, and isolated
the 279 nonpharyngeal neurons, with a total of 6393 chemical synapses
and 890 gap junctions originally cleaned up in \autocite{Chen2006}. A
cell's position was its distance along the anterior-posterior axis
normalzed between zero and one. We used both networks, the chemical
network as a directed graph and the electrical network as undirected
graph. We use the synapse counts with the logistic-distance poisson
likelihood, scaling the counts by 4.0 to compensate for 
the Poisson's overdispersion.

\subsection{Microprocessor}
We extracted the connection graph for the transistors in the MOS6502
\autocite{visual6502source}. Each transistor has three terminals (gate,
source, drain), but the methods of the original dataset were unable to
consistently resolve which of the C1 and C2 terminals were source and
drain, leading to ambiguity in our encoding. We identified a region
consisting of consisting of three registers X, Y, and S via visual
inspection and focused our efforts there. We created a total of six
connectivity matrices by examining possible terminal parings. One
graph, for example, $R^{gc_1}(e_i, e_j)=1$ if transistor $e_j$ and
$e_j$ are connected via pins $g$ and $c_1$.

@@

@@

\subsection{Hyperprior grids}
\label{supp:hyperpriors}

@@

\todo{Explain Adjusted Rand Index}

>>>>

%\documentclass{nature}
\documentclass{article}
\usepackage[english,american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{todonotes}
\usepackage{subfiles}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

% \begin{affiliations}
%   \item Something Something
%   \item Something Something
% \end{affiliations}

\begin{abstract}
  New techniques produce massive data about neural connectivity,
  necessitating new analysis methods to discover underlying
  structure. It has long been assumed that discovering the local patterns of microcircuitry is the key to understanding neural function. Here we developed a
  nonparametric Bayesian technique that identifies neuron types and circuitry
  patterns in connectomics data. We show that the approach recovers
  known neuron types in the retina, reveals interesting structure in
  the nervous system of c. elegans, and automatically discovers the
  structure of microprocessors. Our approach extracts structural meaning from connectomics, enabling new approaches of deriving anatomical insights from these emerging datasets.

\end{abstract}

@@

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs,
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development process.

@@

\begin{figure}
  \centering 
    \centerline{\includegraphics[width=183mm]{overview.ai}}
  \caption{An overview of our method. a. Input data is
an adjacency matrix indicating whether cell $i$ (row)
synapses on cell $j$ (column). The location of each cell in space is known. 
b. Our algorithm discovers hidden cell types in this connectivity data
by assuming all cells of a type share a distance-dependent connectivity 
profile, c.), with cells of other types. This results in a clustering 
of the cells by those hidden types. b.) shows the cell adjacency
matrix with cells of the same type grouped together. d.) shows
the learned probability of connection between our different types
at various distances, illustrating short-range and long-range
connectivity patterns.}
\label{fig:overview}
\end{figure}

\section{Results}
We use a Bayesian structure discovery approach that is based on
assumptions about the way positions and cell types affect the
probability of connections between cells, a so-called generative
model. The probability of connections between pairs of neurons
decreases with distance, but how high the baseline probability of
connection is and how quickly it decays with distance depends on the
two cell-types. We combine annealing with Markov Chain Monte Carlo to
perform simultaneously joint posterior inference over the structure,
parameters, and hyperparameters of the model (see Methods). This
procedure automatically learns the number of cell types, which cells 
belong to which types, along with the 
the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). These results can then be compared with the
results of previous anatomical studies that have used other aspects of
neurons, such as chemical properties and cell-morphology to divide
neurons into types.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus started out by
simulating data for which we know the correct structure and comparing
the estimated structure based on the algorithm (see methods) with the
one we used for simulation. We find that the model does a good job of
recovering the structure we used to generate the data
(fig~\ref{fig:synthetic}a) Indeed, it does far better than a simpler
and well established model (the infinite stochastic block model, or
ISBM) that assumes that there is no distance dependence to the
connectivity. The model converges relatively quickly to an estimate of
the most probable values for the cell types, which is enabled by using
a combination of simulated annealing and parallelized Markov chain
Monte Carlo (see methods for details). It appears that the model
behaves the way it should and allows application to relatively large
datasets.

\begin{figure}
  \centering 
  \centerline{\includegraphics{synthetic.ai}}
  \caption{Correct recovery of true numbers of hidden classes
in synthetic data. a.) Other models over-cluster the number
of classes as they fail to take distance into account, whereas
our model finds close to the true number of classes. b.) ARI, a measure
of ``correct clustering'', for
different true class counts, between our model and a traditional
ISBM. c.) For synthetic data, 
the IRM over clusters, finding a large number of spatially-localized types. 
In contrast, the spatial-relational model recovers the true spatial
extent of the underlying types.}
\label{fig:synthetic}
\end{figure}

@@

The algorithm took 8 hours to converge to a locally optimal solution,
dividing neurons into a set of cell types (fig~\ref{fig:mouseretina}c, each
wedge is a type). For each pair of neurons there is a specific
distance dependent connection probability (fig~\ref{fig:mouseretina}b,c,d),
which is well approximated by the model fit. Moreover, each type of
cell is rather isotropically distributed across space
(fig~\ref{fig:mouseretina}e) as should be expected for true cell types. The
algorithm is thus able to extract meaningful structure from the data.

Comparing the results of the algorithm to other information sources
allows evaluating the quality of the type determination. Our types
closely reflect the (anatomist-determined) segmentation of cells into
retinal ganglion, narrow amacrine, medium/wide amacrine, and bipolar
cells (fig~\ref{fig:mouseretina}c, outermost ring). We find that the
types we find tend to reflect the known laminar distribution in the
retina (fig~\ref{fig:mouseretina}c, middle ring), which is exciting as
the algorithm did not actually use this type of information.

The algorithm yields a separation of neurons into a smaller number of
types than the fully granular listing of 71 types found by the
original authors of the paper, although is still highly correlated
with those finer type distinctions (see
supp~\ref{supp:mouseretina}. It is our expectation that, with larger
datasets, even closer agreement would be found.

\todo{Do we assign any of the unknown cells correctly?  are types 72-79
“different” or simply unknown?}

We compared the results of our model to the traditional ISBM to see if
the addition of spatial information was necessary. We find that, as expected from our synthetic experiments, the
IRM massively over-segmented the data, finding
spatially-localized “communities” of cells which bore little relation
to known types.  The average spatial extent of the types found by the ISBM
was  $14.6\pm8.1 \mu m$. However, in the ground truth by
anatomists the average distance was $34.3\pm7.6 \mu m$ and $35.3\pm 2.4 \mu m$, quite
close to the $34.5\pm 1.0 \mu m$ discovered by our algorithm. Modeling the
distance-dependence prevents over-segmentation and thus allows more
meaningful estimates.

% from experiments/mouseretina/spatial_var.txt

% Relational Model: mean = 14.586508 std=8.136334
% Spatial-Relational Model: mean = 34.540355 std=1.069483
% Truth (fine): mean = 34.348443 std=7.567564
% Truth (coarse): mean = 35.276636 std=2.372951

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome. 
a.) Input adjacency data for 950 cells for which soma positions were known. b. clustered adjacency matrix. c.) connectivity diagram showing our clusters, as
well as the cell depth and known cell type of each cell. Our model agrees
with the cell types traditionally identified by anatomists, and identifies 
laminar-specific connectivity patterns. d.) Connectivity between our
clusters as a function of distance -- the cluster consisting primarily of
retinal ganglion cells (lower left-center) exhibits the expected near and
far connectivity.}
\label{fig:mouseretina}
\end{figure}

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons deterministically known. This
incredible cross-organism consistency led to early mapping of the
c. elegans chemical and electrical (gap junction) connectome. Unlike
the retina, only the motor neurons in c. elegans exhibit regular
distribution in space, being distributed along the body axis. Most
interneurons are concentrated in various ganglia that project
throughout the entire animal, and the sensory neurons are primarily
located in a small number of anterior ganglia.

@@

We identify cell types that reflect the known motor/non-motor neuron
classes, even though this system lacks the strong repeat
microcircuitry our model was designed for.  Motor-neuron types AS, DA,
and VA, all exclusively postsynaptic, are clustered together, as are
motor-neuron types VD and DD. Traditional types VC, DB, and VB also
mostly share a cluster. Various head motor neurons, including types
SMD, RMD, and RIM are clustered together. Interneurons with known
anatomically-distinct connectivity patterns, such as RIA (2 cells) or
AVA (2 cells) are clustered into pure types. The algorithm even
correctly places the single-cell type DVB by itself. Note our
clustering does not perfectly reflect known divisions -- several
combinations of head and sensory neurons are combined, and a
difficult-to-explain group of 2 VD, 2 VB, a VA, and a DD neuron are
grouped together. Our identified cell types thus reflect a
``coarsening'' of known types, based entirely on connectivity and
distance information, even when the organism exhibits substantially
less spatial regularity than the retina.

\begin{figure}
  \centering 
  \centerline{\includegraphics{celegans.ai}}
  \caption{Discovering cell types in the \textit{c. elegans}
    connectome. a.) Initial cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). }
  \label{fig:celegans}
\end{figure}

@@

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:microprocessor}a, showing the man-made horizontal and
vertical layout of the same types of transistors.

@@

  Larger datasets will generally allow algorithms to distinguish more
  distinct types (cite supplemental data) and more data should
  algorithms to get closer to the results from anatomy. Moreover, in
  general, for such problems precision increases with the size of the
  dataset and the cells that we have are not sufficient to
  statistically distinguish all the cell types known in anatomy
  ($\sim 70$). Still, using only connectivity and distance it is possible
  to meaningfully divide neurons into types.

@@

There have existed a range of previous approaches to the discovery
of neural microcircuitry \autocite{} These generally involve a great deal
of manual labor and ad-hoc determination of what constitutes a “type”
of cell -- to this day there are disagreements in the literature as to
the “true” types in the mammalian retina. Much as phylogenomics has
changed our understanding of animal ontologies, modern large scale
data will allow the efficient unbiased discovery of cell types and
circuits. The sheer amount of available data demands the introduction
of algorithmic approaches.

The development of automatic identification and quantification of cell
type may also provide a new computational phenotype for quantifying the effect
of disease, genetic interventions, and developmentally-experienced neural activity. Our method can in principle identify neuron-types across non-connected graphs, e.g. across animals. For example, the types of neurons in one animal can be associated with the types of neurons in another animal, in the same way as this is already possible through molecular markers \autocite{}. This could be particularly important if cell types appear that are due to properties of the stimuli and experience as opposed to the molecular properties of cells \autocite{}. This would allow comparative quantitative anatomy across animals, and the search for the causes of the ultimate connectivity.

@@

We then jointly infer the maximum a posteriori (MAP) estimate of the
class assignment vector ${c_i}$, the parameter matrix $\eta_{mn}$, and
the global model hyperparameters $\theta$ :

@@

% Methods section 3000 words
% methods summary does not appear in the online text
% methods section can have no figures or tables [grrr]

@@

Our model is a extension of the ISBM
\autocite{Kemp2006a,Xu2006} to incorporate spatial relations between entities,
inspired by attempts to extend these models with arbitrary
discriminative functions\autocite{Murphy2012}.

@@

We then employ a transition kernel to update the per-component
parameter values $\eta_{mn}$. Conditioned on the assignment vector
$\vec{c}$ and the model hyperparameters $\theta, \alpha$ the 
individual $\eta_{mn}$ are independent. We slice sample \autocite{Neal2003}
each component's parameters, choosing the slice width as a function
of the global hyperparameter range.

@@

To pick the MAP, we run up to $100$ simultaneous chains in parallel,
initialized from different random initial points in the state space,
and then pick the chain with the highest log likelihood. Early
experiments attempted to use the full ensemble of chains as an
estimate of the posterior but were abandoned due to the difficulty of
summarizing the posterior over such a complex state space, which is an
ongoing area of research in the probabilistic modeling community
\autocite{}.

@@

\subsection {Mouse Retina}
Dense serial electron microscopy of a $VOLUMNE$ in the mouse retina by
\autocite{Helmstaedter2013} yielded a listing of places where neurons
come into contact. There were over 1000 cells originally, and selected
the $950$ for which the location of the soma could be reconstructed
from the provided cell plots (soma locations were not provided by the
study's authors in machine-readable form). Ultimately
this left a matrix between the total synapse-like contact area between
all pairs of 950 cells. Area was thresholded at $0.5\mu m$, determined
by hand, to yield a 950 $\times$ 950 entry matrix that served as input
to our algorithm. We measured the distance between cells using
the reconstructed soma centers, and used the Logistic-Distance
spatial relation. Hyperprior griddings are shown in \ref{supplemental}.

@@

We obtained the connectome of c. elegans from \autocite, and isolated
the 279 nonpharyngeal neurons, with a total of 6393 chemical synapses
and 890 gap junctions originally cleaned up in \autocite{Chen2006}. A
cell's position was its distance along the anterior-posterior axis
normalized between zero and one. We used both networks, the chemical
network as a directed graph and the electrical network as undirected
graph. We use the synapse counts with the logistic-distance poisson
likelihood, scaling the counts by 4.0 to compensate for 
the Poisson's overdispersion.

\subsection{Microprocessor}
We extracted the connection graph for the transistors in the MOS6502
\autocite{visual6502source}. Each transistor has three terminals (gate,
source, drain), but the methods of the original dataset were unable to
consistently resolve which of the C1 and C2 terminals were source and
drain, leading to ambiguity in our encoding. We identified a region
consisting of consisting of three registers X, Y, and S via visual
inspection and focused our efforts there. We created a total of six
connectivity matrices by examining possible terminal pairings. One
graph, for example, $R^{gc_1}(e_i, e_j)=1$ if transistor $e_j$ and
$e_j$ are connected via pins $g$ and $c_1$.

@@

\FloatBarrier

@@

\FloatBarrier
\subsection{Comparison to IRM for Mouse Retina}

\begin{figure}[h!]
  \centering 
    \centerline{\includegraphics[width=5in]{mouseretina/spatial_var.pdf}}
  \caption{}
\label{fig:supp:irmspatialcomparemouse}
\end{figure}

\begin{figure}[h!]
  \centering 
    \centerline{\includegraphics[width=6.5in]{mouseretina/retina.1.2.bb.0.0.data-fixed_20_100-anneal_slow_400.truth_latent.pdf}}
  \caption{}
\label{fig:supp:retinabbclusters}
\end{figure}

\begin{figure}[h!]

  \centering 
    \centerline{\includegraphics[width=6.5in]{mouseretina/retina.1.2.ld.0.0.data-fixed_20_100-anneal_slow_400.truth_latent.pdf}}
  \caption{}
\label{fig:supp:retinaldclusters}
\end{figure}

\newpage

\FloatBarrier
\subsection{Hyperprior grids}
\label{supp:hyperpriors}

@@

\section{Measuring clustering similarity}

The adjusted rand index (ARI) is a measure of the similarity of two different clusterings \autocite{Rand1971} -- two identical clusters have an ARI of 1.0 while progressively more dissimilar clusters have lower ARIs, becoming negative as the clustering gets anti-correlated.
